---
title: "Future constructions in English and Norwegian"
author: "Olaf Mikkelsen & Stefan Hartmann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This file contains the scripts used for a contrastive analysis of English and Norwegian future constructions.

# Preliminaries

Load packages

```{r}

library(tidyverse)
library(readxl)
library(party)
library(lattice)
library(Hmisc)
library(pdp)
library(collostructions) # available at sfla.ch
library(patchwork)
library(ggparty)
library(Boruta)
library(svglite)

```

## Queries

The English data were queried via [CWB](https://cwb.sourceforge.io/). SpokenBNC2014 was transformed to CWB format using [this](https://github.com/hartmast/SpokenBNC2014_to_CWB) script. OANC was transformed to CWB using [this](https://github.com/hartmast/OANC_to_CWB) script. Uncollapse the following code box to see the queries. 


```{r class.source = 'fold-hide', eval=FALSE}

# Spoken BNC 2014
cqp -e
SPOKENBNC2014
[word="will|shall|going|wo|'?ll|gon"][word="to|n't|na"]?
cat Last > "will_shall_going_to_SPOKENBNC2014.txt"

# in addition, for taking priming effects into account:
cqp -e
SPOKENBNC2014;
set context 150 words;
show +u_who;
set ld "üüü";
set rd "üüü";
[word="will|shall|going|wo|'?ll|gon"][word="to|n't|na"]?;
cat Last > will_shall_going_to_SPOKENBNC2014_more_context.txt

In the text file, replace üüü by \t.

# OANC
cqp -e
OANCSPOKEN
set context s
show +lemma +pos
set PrintStructures "text_id, text_genre, text_file, turn_id, turn_age, turn_sex"
A = [word="will|shall|going|wo|ll|gon|gonna"][word="to|n't|na"]?
cat A > "oanc_spoken.txt"

# additional query for 'll files because the apostrophe
# was missing in the original query
cqp -e
OANCSPOKEN
set context s
show +lemma +pos
set PrintStructures "text_id, text_genre, text_file, turn_id, turn_age, turn_sex"
A = [word="'ll"]
cat A > "oanc_ll.txt"

```

# Data wrangling

## Read data

```{r}

# read data: Norwegian
nor_bb <- read_xlsx("data/Norwegian/nota_bb_vil_skal_komer_til_a.xlsx", sheet = "bigbrother")
nor_nota <- read_xlsx("data/Norwegian/nota_bb_vil_skal_komer_til_a.xlsx", sheet = "nota")
nor <- rbind(nor_nota, nor_bb)

# clean column names
colnames(nor) <- gsub(" ", "_", colnames(nor))

# read data: English
oanc <- read_csv("data/English/OANC/oanc_spoken_going_to_shall_will_5000.csv")
bnc  <- read_xlsx("data/English/SpokenBNC2014/SPOKENBNC2014_spoken_will_shall_going_to_sample_5000-final-priming.xlsx")

```

## Add annotation columns

Protasis/apodosis annotation columns are added automatically.

```{r}

# add protasis/apodosis annotation to BNC
bnc$if_clause2 <- ifelse(bnc$if_clause=="if" & bnc$subordinate=="sub", "protasis", NA)
bnc$if_clause2 <- ifelse(bnc$if_clause=="if" & bnc$subordinate=="main", "apodosis", bnc$if_clause2)
bnc$if_clause2 <- ifelse(bnc$if_clause!="if", "no", bnc$if_clause2)

# add protasis/apodosis annotation to OANC
oanc$if_clause2 <- ifelse(oanc$if_clause=="if" & oanc$subordinate=="sub", "protasis", NA)
oanc$if_clause2 <- ifelse(oanc$if_clause=="if" & oanc$subordinate=="main", "apodosis", oanc$if_clause2)
oanc$if_clause2 <- ifelse(oanc$if_clause!="if", "no", oanc$if_clause2)

# add protasis/apodosis annotation to NOR
nor$if_clause2 <- ifelse(nor$`If-clause`=="if" & nor$Clause_type=="sub", "protasis", "NA")
nor$if_clause2 <- ifelse(nor$`If-clause`=="if" & nor$Clause_type=="main", "apodosis", nor$if_clause2)
nor$if_clause2 <- ifelse(nor$`If-clause`!="if", "no", nor$if_clause2)

```

Also, we add columns with the outcome variable to the English data, with an additional column with a binary encoding (will / going to)

```{r}

# OANC ---------

# binary:
oanc$cxn <- ifelse(oanc$Key %in% c("gonna", "going"), "going_to", "will")

# more fine-grained:
oanc$cxn1 <- case_when(oanc$Key == "will" ~ "will",
                       oanc$Key == "'ll" ~ "'ll",
          oanc$Key == "shall" ~ "shall",
          oanc$Key == "wo" ~ "won't",
          oanc$Key == "going" ~ "going to",
          oanc$Key == "gonna" ~ "gonna")
oanc$cxn1 <- factor(oanc$cxn1, levels = c("will", "'ll", "shall", "won't", "going to", "gonna"))

# BNC ---------

# binary
bnc$cxn <- ifelse(bnc$Key %in% c("'ll", "will", "wo"), "will", "going to")

bnc$cxn1 <- case_when(bnc$Key == "gon" ~ "gonna",
          bnc$Key == "'ll" ~ "'ll",
          bnc$Key == "going" ~ "going to",
          bnc$Key == "will" ~ "will",
          bnc$Key == "wo" ~ "won't",
          bnc$Key == "shall" ~ "shall")

bnc$cxn1 <- factor(bnc$cxn1, levels = c("will", "'ll", "shall", "won't", "going to", "gonna"))

```

Omit false hits:

```{r}
bnc  <- bnc %>% filter(keep == "y")
oanc <- oanc %>% filter(keep=="y")
```


# Summary statistics

Number of hits per corpus:

```{r}

left_join(setNames(as.data.frame(table(bnc$cxn1)), c("Cxn", "BNC")), 
          setNames(as.data.frame(table(oanc$cxn1)), c("Cxn", "OANC")))

```


# CART Tree & Random forest - Norwegian

Fitting a conditional inference tree and a random forest to the Norwegian data:

```{r}

# Norwegian: ---------------

# relevant variables as factors
nor$Cx <- factor(nor$Cx)
nor$Negative <- factor(nor$Negative)
nor$Interrogative <- factor(nor$Interrogative)
nor$if_clause2 <- factor(nor$if_clause2)
nor$Clause_type <- factor(nor$Clause_type)

# reduce number of levels for Lexeme
nor$Lexeme <- factor(nor$Lexeme)
nor$lexeme <- fct_lump_min(nor$Lexeme, min = 50, other_level = "other")


# more descriptive name
nor$Construction <- nor$Cx

# shorter cx labels (obsolete now that we're using ggparty)
nor$cxn <- case_when(nor$Cx == "kommer" ~ "k",
          nor$Cx == "skal" ~ "s",
          nor$Cx == "vil" ~ "v")
nor$cxn <- factor(nor$cxn)

# shorter if clause label
nor$if_clause <- nor$if_clause2

# CART tree
set.seed(19851003)
tr_nor <- ctree(Construction ~ Negative+Interrogative+if_clause+Clause_type,
                data = nor)

# plot - adapted from https://ladal.edu.au/tree.html

# extract p-values
pvals <- unlist(nodeapply(tr_nor, ids = nodeids(tr_nor), function(n) info_node(n)$p.value))
pvals <- pvals[pvals <.05]

( tr_nor_plot <- ggparty(tr_nor) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(line_list = list(aes(label = splitvar),
                                   aes(label = paste0("N=", nodesize, ", p", 
                                                      ifelse(pvals < .001, "<.001", paste0("=", round(pvals, 3)))), 
                                       size = 10)),
                  line_gpar = list(list(size = 13), 
                                   list(size = 10)), 
                  ids = "inner") +
  geom_node_label(aes(label = paste0("Node ", id, ", N = ", nodesize)),
    ids = "terminal", nudge_y = -0.0, nudge_x = 0.01) +
  geom_node_plot(gglist = list(
    geom_bar(aes(x = "", fill = Construction),
             position = position_fill(), color = "black"),
    theme_minimal(),
    scale_fill_grey(start = .4, end = .9),
    scale_y_continuous(breaks = c(0, 1)),
    xlab(""), 
    ylab("Probability"),
    geom_text(aes(x = "", group = Construction,
                  label = after_stat(count)),
              stat = "count", position = position_fill(), vjust = 1.1)),
    shared_axis_labels = TRUE) )
    
# export as pdf (ggsave not working for ggparty objects apparently)

# png("figures/tree_NO.png", width = 10, height = 6, un = "in", res = 300)
# plot(tr_nor)
# dev.off()


```


```{r, eval = FALSE}

# random forest
set.seed(19851003)
for_nor <- party::cforest(Cx ~ Negative+Interrogative+if_clause+Clause_type,
                data = nor, 
                controls = cforest_unbiased(mtry = 2, ntree = 2000))
vi <- party::varimp(for_nor, conditional = TRUE)


```

```{r, include = FALSE, eval = FALSE}

# export random forest
write_rds(for_nor, "rds/for_nor.rds")

# export varimp
write_rds(vi, "rds/vi.rds")

```

```{r, include = FALSE}

# re-import random forest
for_nor <- read_rds("rds/for_nor.rds")

# re-import variable importance
vi      <- read_rds("rds/vi.rds")

```

Plot the variable importance:

```{r}


(vi_plot_nor <- as.data.frame(vi) %>% rownames_to_column() %>% ggplot(aes(y = fct_reorder(rowname, vi), x = vi)) + 
  geom_point(size = 5) + xlim(min(vi), max(vi)) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  theme(panel.grid.minor = element_blank()) + xlab("Conditional variable importance") + ylab("Variable") +
  theme(axis.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18)) +
  theme(strip.text = element_text(size = 18)) +
  theme(legend.text = element_text(size = 18)) +
  theme(legend.title = element_text(size = 18, face = "bold")) +
  theme(text = element_text(size = 18)))


```

## Check model fit

Model accuracy

```{r, eval = FALSE}

# fitted values for out of bag sample
pred.nor <- predict(for_nor, OOB = T)


```


```{r, include = FALSE, eval = FALSE}

# export fitted values
write_rds(pred.nor, "rds/pred_nor.rds")


```


```{r, include = FALSE}

# re-import fitted values
pred.nor <- read_rds("rds/pred_nor.rds")

```


```{r}

# proportion of correct predictions
sum(as.numeric(sapply(1:length(pred.nor), function(i) pred.nor[i] == nor$Cx[i]))) / length(pred.nor)

```


## CART Tree and Random Forest - British and American English

```{r}

# combine OANC and BNC2014 data

colnames(oanc)[which(colnames(oanc)=="gramm_person")] <- "Gramm_person"
colnames(oanc)[which(colnames(oanc)=="animacy")] <- "Animacy"
colnames(oanc)[which(colnames(oanc)=="question")] <- "Interrogative"
colnames(bnc)[which(colnames(bnc)=="question")] <- "Interrogative"
colnames(bnc)[which(colnames(bnc)=="subordinate")] <- "Clause_type"
colnames(oanc)[which(colnames(oanc)=="subordinate")] <- "Clause_type"
colnames(oanc)[which(colnames(oanc)=="negation")] <- "Negative"
colnames(bnc)[which(colnames(bnc)=="negation")] <- "Negative"


eng <- rbind(
  mutate(select(oanc, Left, Key, Right, Clause_type, Gramm_person, Animacy, Interrogative, Negative, if_clause, if_clause2, cxn, cxn1, Lemma), corpus = "OANC-Spoken"),
mutate(select(bnc, Left, Key, Right,  Clause_type, Gramm_person, Animacy, Interrogative, Negative, if_clause, if_clause2, cxn, cxn1, Lemma), corpus = "SpokenBNC2014")
)

# CART tree
eng$Negative <- factor(eng$Negative)
eng$Interrogative <- factor(eng$Interrogative)
eng$if_clause2 <- factor(eng$if_clause2)
eng$Clause_type <- factor(eng$Clause_type)
eng$cxn <- factor(eng$cxn)
eng$cxn1 <- factor(eng$cxn1)


# omit NAs
eng1 <- select(eng, Negative, Interrogative, if_clause2, Clause_type, cxn1, corpus, Lemma)
eng1 <- na.omit(eng1)

# abbreviate labels - obsolete using ggparty
eng1$cxn <- case_when(eng1$cxn1=="will" ~ "w",
          eng1$cxn1=="shall" ~ "s",
          eng1$cxn1=="won't" ~ "wnt",
          eng1$cxn1=="going to" ~ "gt",
          eng1$cxn1=="gonna" ~ "gna",
          eng1$cxn1=="'ll" ~ "ll")

# add lexeme
eng1$Lexeme <- factor(eng1$Lemma)

# add lexeme with reduced number of levels (bin infrequent lemmas to "other")
eng1$lexeme <- fct_lump_min(eng1$Lexeme, min = 35)


# factor
eng1$Construction <- eng1$cxn1
eng1$cxn <- factor(eng1$cxn, levels = c("w", "ll", "wnt", "s", "gt", "gna"))

# rename if-clause column to increase readability of tree diagram
eng1$if_clause <- eng1$if_clause2

# corpus as factor
eng1$corpus <- factor(eng1$corpus)

# more descriptive: use language variety instead of corpus
eng1$Variety <- ifelse(eng1$corpus == "OANC-Spoken", "AmE", "BrE")
eng1$Variety <- factor(eng1$Variety)

set.seed(1985)
tree_en <- ctree(Construction ~ Negative+Interrogative+if_clause+Clause_type+Variety,
                data = eng1)


# plot

# extract p-values
pvals <- unlist(nodeapply(tree_en, ids = nodeids(tree_en), function(n) info_node(n)$p.value))
pvals <- pvals[pvals <.05]

# plot:
( tr_en_plot <- ggparty(tree_en) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(line_list = list(aes(label = splitvar),
                                   aes(label = paste0("N=", nodesize, ", p", 
                                                      ifelse(pvals < .001, "<.001", paste0("=", round(pvals, 3)))), 
                                       size = 10)),
                  line_gpar = list(list(size = 13), 
                                   list(size = 10)), 
                  ids = "inner") +
  geom_node_label(aes(label = paste0("Node ", id, ", N = ", nodesize)),
    ids = "terminal", nudge_y = -0.0, nudge_x = 0.01) +
  geom_node_plot(gglist = list(
    geom_bar(aes(x = "", fill = Construction),
             position = position_fill(), color = "black"),
    theme_minimal(),
    scale_fill_grey(start = .4, end = .9),
    scale_y_continuous(breaks = c(0, 1)),
    xlab(""), 
    ylab("Probability"),
    geom_text(aes(x = "", group = Construction,
                  label = after_stat(count)),
              stat = "count", position = position_fill(), vjust = 1.1)),
    shared_axis_labels = TRUE) )

# save as pdf via export pane

dev.off()
# png("tree_en.png", width = 30, height = 10, un = "in", res = 300)
plot(tree_en)
# dev.off()

# random forest

```

```{r, eval=FALSE}
set.seed(1985)
for_en <- party::cforest(Construction ~ Negative+Interrogative+if_clause+Clause_type+Variety,
                data = eng1, controls = cforest_unbiased(mtry = 3, ntree = 2000))
vi_en <- party::varimp(for_en, conditional = TRUE)

```

```{r, include=FALSE, eval=FALSE}
# export random forest and vi

# export random forest
write_rds(for_en, "rds/for_en.rds")

# export varimp
write_rds(vi_en, "rds/vi_en.rds")

```

```{r, include = FALSE}

# re-import saved random forest
for_en <- read_rds("rds/for_en.rds")

# re-import 
vi_en <- read_rds("rds/vi_en.rds")

```



Visualize variable importance:

```{r}


(vi_plot_en <- as.data.frame(vi_en) %>% rownames_to_column() %>% ggplot(aes(y = fct_reorder(rowname, vi_en), x = vi_en)) + 
  geom_point(size = 5) + xlim(min(vi_en), max(vi_en)) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  theme(panel.grid.minor = element_blank()) + xlab("Conditional variable importance") + ylab("Variable") +
  theme(axis.text = element_text(size = 18)) +
  theme(axis.title = element_text(size = 18)) +
  theme(strip.text = element_text(size = 18)) +
  theme(legend.text = element_text(size = 18)) +
  theme(legend.title = element_text(size = 18, face = "bold")) +
  theme(text = element_text(size = 18)))

# both plots
vi_plot_nor + ggtitle("Norwegian") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) + 
  plot_spacer() +
  vi_plot_en + ggtitle("English") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  plot_layout(widths = c(4,.5,4))
# ggsave("figures/forest_NOR_EN.png", width = 12, height = 5)


```


## Check model fit

Model accuracy (number of correct predictions / number of observations)

```{r, eval = FALSE}

# fitted values for out of bag sample
pred.eng <- predict(for_en, OOB = T)

```


```{r, include = FALSE, eval = FALSE}

# export fitted values
write_rds(pred.eng, "rds/pred_eng.rds")

```


```{r, include = FALSE}

# re-import fitted values
pred.eng <- read_rds("rds/pred_eng.rds")

```


```{r}

table(pred.eng)
table(eng$cxn1)
table(pred.eng, eng1$cxn1)
sum(as.numeric(sapply(1:length(pred.eng), function(i) pred.eng[i] == eng1$cxn1[i]))) / length(pred.eng)


```


# Boruta analysis

## Norwegian

```{r}

# first Boruta model for Norwegian
nor$Lexeme <- factor(nor$Lexeme)
set.seed(19551105)
boruta01 <- Boruta(Cx ~ Negative+Interrogative+if_clause+Clause_type+Lexeme,
                data = nor)

# decision
getConfirmedFormula(boruta01)
plotImpHistory(boruta01)
plot(boruta01)

# Boruta model with confirmed formula
set.seed(19551105)
boruta02 <- Boruta(Cx ~ Lexeme + Clause_type + if_clause +
                     Interrogative + Negative,
                   data = nor)


(boruta_NOR <- as.data.frame(boruta02$ImpHistory) %>% pivot_longer(cols = 1:length(as.data.frame(boruta02$ImpHistory) )) %>% setNames(c("Variable", "Importance")) %>%
  mutate(Type = ifelse(str_detect(Variable, "shadow"), "Control", "Predictor")) %>%
  mutate(Type = factor(Type),
                Variable = factor(Variable)) %>%
  ggplot(aes(x = reorder(Variable, Importance, mean), y = Importance, fill = Type)) + 
  geom_boxplot() +
  theme_bw() +
  scale_fill_grey(start = .5, end = .9) +
  xlab("Variable") + ylab("Importance") + ggtitle("Norwegian") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  theme(panel.grid.major.y = element_line(linewidth = .2, color = "grey75"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()))


```


## English

```{r}

# first Boruta model for English
set.seed(2015)
boruta_en01 <- Boruta(cxn ~ Negative+Interrogative+if_clause+Clause_type+corpus+Lexeme,
                data = eng1)


# decision
getConfirmedFormula(boruta_en01)
plotImpHistory(boruta_en01)

# Boruta model with confirmed formula
boruta_en02 <- Boruta(cxn ~ Lexeme + corpus + Clause_type + if_clause + Interrogative + Negative,
                      data = eng1)

(boruta_EN <- as.data.frame(boruta_en02$ImpHistory) %>% pivot_longer(cols = 1:length(as.data.frame(boruta_en02$ImpHistory))) %>% setNames(c("Variable", "Importance")) %>%
  mutate(Type = ifelse(str_detect(Variable, "shadow"), "Control", "Predictor")) %>%
  mutate(Type = factor(Type),
                Variable = factor(Variable)) %>%
  ggplot(aes(x = reorder(Variable, Importance, mean), y = Importance, fill = Type)) + 
  geom_boxplot() +
  theme_bw() +
  scale_fill_grey(start = .5, end = .9) +
  xlab("Variable") + ylab("Importance") + ggtitle("English") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  theme(panel.grid.major.y = element_line(linewidth = .2, color = "grey75"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank()))


# both plots
boruta_NOR +
  theme(axis.text.x = element_text(angle=45, hjust=.9, size=12)) +
  plot_spacer() + boruta_EN +
  theme(axis.text.x = element_text(angle=45, hjust=.9, size=12)) +
  plot_layout(guides = "collect", widths = c(4,1,4))
# ggsave("figures/boruta.png", width = 15, height = 6.5)

```


# Collostructional analysis

(Multiple) distinctive collexeme analysis is used to find the lexical items that are most strongly associated with the three variants.


## Covarying-collexeme analysis: Norwegian

```{r}

mdca_nor <- nor %>% select(Cx, Lexeme) %>% as.data.frame %>% collex.covar()

mdca_nor %>% filter(SLOT1=="vil")
mdca_nor %>% filter(SLOT1=="skal")
mdca_nor %>% filter(SLOT1=="kommer")


```


## Covarying-collexeme analysis: British English

```{r}

mdca_bnc <- bnc %>% select(cxn1, Lemma) %>% as.data.frame() %>% collex.covar()

mdca_bnc %>% filter(SLOT1=="going to")
mdca_bnc %>% filter(SLOT1=="gonna")
mdca_bnc %>% filter(SLOT1=="will")
mdca_bnc %>% filter(SLOT1=="won't")
mdca_bnc %>% filter(SLOT1=="shall")
```

## Covarying-collexeme analysis: American English

```{r}

mdca_oanc <- oanc %>% select(cxn1, Lemma) %>% as.data.frame() %>% collex.covar()

mdca_oanc %>% filter(SLOT1=="going to")
mdca_oanc %>% filter(SLOT1=="gonna")
mdca_oanc %>% filter(SLOT1=="will")
mdca_oanc %>% filter(SLOT1=="won't")
mdca_oanc %>% filter(SLOT1=="shall")


```


## Lemma lists

The following CWB commands give the lemma lists for SpokenBNC2014 and the spoken component of OANC:

```{zsh, eval=FALSE}
cwb-scan-corpus spokenbnc2014 lemma > "spokenbnc2014_lemmas.tsv"
cwb-scan-corpus oancspoken lemma > "oancspoken_lemmas.tsv"
```

The lemma lists can be used to conduct simple collexeme analyses for the individual constructions.


